---
title: "Halloween_Candy_lab: Spooky Bioinformatics!"
author: "Monika Ramos"
format: html
---

**Exploratory Analysis of Halloween Candy**
```{r}
candy_file <- "candy.csv"
```
```{r}
candy = read.csv(candy_file, row.names=1)
head(candy)
```
Q1. How many different candy types are in this dataset?
```{r}
dim(candy)
```
There are 85 candy types and 12 candy groups/measurements in the columns. 

Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```
There are 38 fruity candy types.

**2. What is your favorate candy?**
One of the most interesting variables in the dataset is winpercent. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

```{r}
candy["Twix", ]$winpercent
```

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["100 Grand",]$winpercent
```
My favorite candy is 100 grand bars because it is a hybrid of crunch bars and twix! Its winpercent value is 66.9%.

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```

Kit Kat winpercent value is 76%.

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
This one is slightly lower than others. Its winpercent is 49.65%


There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library(skimr)
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
The histogram is the only column when viewing with skim() function that is a character value rather than integers or doubles with the other columns. Some columns such as chocolate are either 1 or 0 and nothing in between. Mean appears to have the widest range of values.

Q7. What do you think a zero and one represent for the candy$chocolate column?

I think it is similar to logical values where 1 is yes/TRUE and 0 is No/False. This means is the candy considered a chocolate type of candy? If yes, 1 and If no, 0. 


Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```


Q9. Is the distribution of winpercent values symmetrical?
The distribution is not symmetrical. In a symmetrical distribution, mean = median = mode and in this case the mean doesn't **exactly** equal the median, though it is close.

Q10. Is the center of the distribution above or below 50%?
```{r}
summary(candy$winpercent)
```
The center of distribution is roughly at 50% given that the median is 47.83 and the mean is 50.32. With this information, i would say it is just below 50%. 

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
```
```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
```

On average, it appears that chocolate candy is ranked higher than fruity candy.

Q12. Is this difference statistically significant?
```{r}
t.test(x = (candy$winpercent[as.logical(candy$chocolate)]), y=(candy$winpercent[as.logical(candy$fruity)]))
```
The difference is statistically significant with a p-value < 0.05. 

**3. Overall Candy Rankings**
use the base R order() function together with head() to sort the whole dataset by winpercent

Q13. What are the five least liked candy types in this set?
```{r}
library(tidyverse)
candy_sorted <- arrange(.data = candy, winpercent)
head(candy_sorted)
```
The five least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters. Ouch, jawbusters are hard to chew!

Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy_sorted)
```
The top 5 all time favorite candies are ReeseOs Peanut butter cup, ReeseOs Miniatures, Twix, Kit Kat, and Snickers.

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, (rownames(candy))) + geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) + 
 aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

Time to add some useful color. setup a color vector. 
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
Q17. What is the worst ranked chocolate candy?
The worst ranked is Sixlets.

Q18. What is the best ranked fruity candy?

The best fruity is starburst.

4. Taking a look at pricepercent
```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses miniatures 

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L Nip.


5 Exploring the correlation structure
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolate and Fruit are anti-correlated
Q23. Similarly, what two variables are most positively correlated?
chocolate and bar. also chocolate and winpercent...

6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale. = T)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)

```

finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus. Yes it makes sense because chocolate is usually individually packaged. Fruity is usally many like sour patch kids and are harder than chocolate since they are hard sugar. 
