---
title: "Mini-Project Investigating Pertussis Resurgence"
author: "Monika Ramos"
format: html
---
Class 18
Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

#Is pertussis on the rise?
```{r}
library(datapasta)

#use datapasta to paste copied CDC data table and paste it as dataframe.
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)
```

Plot the data
```{r}
library(ggplot2)
ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases per year in the USA")

baseplot <- ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases per year in the USA")
```

Two types of pertussis vaccines have been developed: whole-cell pertussis (wP) and acellular pertussis (aP). The first vaccines were composed of ‘whole cell’ (wP) inactivated bacteria. The latter aP vaccines use purified antigens of the bacteria (the most important pertussis components for our immune system). These aP vaccines were developed to have less side effects than the older wP vaccines and are now the only form administered in the United States.

Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?
```{r}
baseplot + 
  geom_vline(xintercept = 1946, color = "red", linetype = 2) +
  geom_vline(xintercept = 1996, color = "blue", linetype = 2)
```
There is a increase or resurgence after year 2000.

Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?
A few years after the introduction of the aP vaccine, there is an evident surge in pertussis cases. I think a few possible explanations for this could be that the aP vaccine may not contain the necessary antigens for effective vaccination against bortadella in comparison to the whole cell vaccine. Additionally, perhaps the pathogen mutated enough to overcome the immunization efficacy and be different enough that the vaccination is no longer efficacious. 


3. Exploring CMI-PB data
Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

The new and ongoing CMI-PB project aims to provide the scientific community with this very information. In particular, CMI-PB tracks and makes freely available long-term humoral and cellular immune response data for a large number of individuals who received either DTwP or DTaP combination vaccines in infancy followed by Tdap booster vaccinations. This includes complete API access to longitudinal RNA-Seq, AB Titer, Olink, and live cell assay results directly from their website: https://www.cmi-pb.org/


```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```
Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

Q5. How many Male and Female subjects/patients are in the dataset?
```{r}
table(subject$biological_sex)
```

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
Side-Note: Working with dates
Two of the columns of subject contain dates in the Year-Month-Day format. Recall from our last mini-project that dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:

```{r}
library(lubridate)
```
```{r}
today()
today() - ymd("2000-01-01")
time_length( today() - ymd("2000-01-01"),  "years")
```

Note that here we are using the ymd() function to tell lubridate the format of our particular date and then the time_length() function to convert days to years.

Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?
```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth),  "years")
#what is the average age of all subjects?
mean(subject$age)
```
```{r}
summary(subject$age)
```
Histogram
```{r}
ggplot(subject) +
  aes(age) +
  geom_histogram()
```
There is an evident peak in younger age groups.
```{r}
#add a faceted wrap layer by wP and aP
ggplot(subject) +
  aes(age) +
  geom_histogram() +
  facet_wrap(vars(infancy_vac))
```
Yes they are different. 


Q8. Determine the age of all individuals at time of boost?
```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?
```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2)
```

Yes they appear very different. 

Joining multiple tables
Read the specimen and ab_titer tables into R and store the data as specimen and titer named data frames.

```{r}
# Complete the API URLs...
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE) 
head(specimen)

```
We want to join the 'subject' and the 'specimen' tables to have all the metadata we need for later analysis. we can use the dplyr *_join() funtions for this task. 
```{r}
library(dplyr)
meta <- inner_join(specimen,subject)
dim(meta)
```
Now read antibody titre experimental data
```{r}
titer <- read_json("http://cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
head(titer)
```
I need to linke a.k.a join the ab titre data with our meta data
```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
This is getting really large. 

4. Examine IgG1 Ab titer levels

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```
Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:
```{r}
ggplot(ig1) +
  aes(MFI, antigen, color= antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r}
table(abdata$isotype)
```


```{r}
table(abdata$visit)
```

4. Examine IgG1 Ab titer levels
```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```
Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:
```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```
Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

It looks like FIM2/3 has the most IgG1 antibody titre and this could be because it is an immunodominant antigen. Perhaps it is more antibody-accessible than the others as well. Lastly, it could be extracellular in contract to intracellular antigens which require MHC-I or MHC-II antigen processing and presentation.


We can attempt to examine differences between wP and aP here by setting color and/or facet values of the plot to include infancy_vac status (see below). However these plots tend to be rather busy and thus hard to interpret easily.



```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```


```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```
Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).
```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

and the same for antigen=="FIM2/3"

```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


